---
title: "pml_final-assignment"
author: "Stefano Galeano"
date: "9/14/2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# PML - Final assignment

## Introduction

## Exploratory data analysis

```{r, message=TRUE, warning=TRUE, include=FALSE}
trainUrl <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
testUrl <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'

download.file(url = trainUrl,destfile = 'data/train.csv')
download.file(url = testUrl,destfile = 'data/test.csv')
train <- read.csv('data/train.csv', header = TRUE, stringsAsFactors = FALSE, na.strings = c("NA","#DIV/0!",""))
test <- read.csv('data/test.csv', header = TRUE, stringsAsFactors = FALSE, na.strings = c("NA","#DIV/0!",""))
```

```{r}
train.sel <- train[train$new_window == "yes",]
test.sel <- test

train.sel$classe <- as.factor(train.sel$classe)

train.sel$X <- NULL
train.sel$user_name <- NULL
train.sel$raw_timestamp_part_1 <- NULL
train.sel$raw_timestamp_part_2 <- NULL
train.sel$cvtd_timestamp <- NULL
train.sel$new_window <- NULL
train.sel$num_window <- NULL
```

```{r}
ncol(train.sel)
naFeatureIndexes <- sapply(names(train.sel),function(featureName){
        sum(is.na(train.sel[,featureName]))/nrow(train.sel) > .8
})
length(which(naFeatureIndexes))
train.sel[,naFeatureIndexes] <- NULL
ncol(train.sel)
```
```{r}
library(Hmisc)

sum(is.na(train.sel))

train.sel[,] <- lapply(train.sel[,],impute,median)

```{r}

sum(is.na(train.sel))
```

```{r}
library(caret)

ntzVars <- nearZeroVar(train.sel)
summary(train.sel[,ntzVars])
train.sel[,ntzVars] <- NULL
```

```{r}
summary(train.sel)
outcomeIndex <- which(names(train.sel) == "classe")
train.cor <- cor(train.sel[,-outcomeIndex])
highlyCorrelated <- findCorrelation(train.cor, cutoff=0.75, names = TRUE)
train.sel[,highlyCorrelated] <- NULL
```

```{r}
```

```{r}
library(caret)

set.seed(1234)

fitTre <- train(classe~., data = train.sel, method = "rpart")
fitTre
```
## Conclusion
